Marginally-Stable Thermal Equilibria of Rayleigh-Bénard Convection

### 2021-04-21 publication_materials/ms_rbc.tex
eigentools cite, broyden's method details

\subsection{Treatment of multiple marginally-stable modes} \label{sec:multiple_modes}
In most cases, we eventually encounter eigenvalue spectra with multiple simultaneously marginal modes.
If ignored, we are forced to reduce the timestep and allow the modes to alternate which is marginally stable.
Instead, we generalize the advective term in \eq{EQ:T0_IVP} to accommodate $N$ simultaneously marginal modes
\begin{equation}
    \langle w' T' \rangle_x = \sum_{n = 1}^{N} 4 A_n^2 \int \, \Re\left[ W_n \theta_n^* \right] \,dz
\end{equation}
where $W_n$ and $\theta_n$ are the eigenmodes associated with $k_x = \frac{n\pi}{2}$. 
There are now $N$ amplitudes to solve for and $N$ eigenvalues to keep marginally-stable. 
Given a small fixed time step $\Delta t$, let $\vec{A^2}$ be the amplitude vector and $\vec{\sigma}(\vec{A^2})$ be the dependent eigenvalues.
We expect a function $\vec{\sigma} \, : \, \mathbb{R}^N \to  \mathbb{R}^N$ to have isolated roots $\vec{A^2_m}$ (should they exist). 
A coarse approximation for the appropriate amplitude vector $\vec{A^2_m}$ is given by a generalized form of \eq{EQ:amp_approx}:
\begin{equation}
    \vec{A^2_m} \approx -\Sigma_{\rm{adv}}^{-1} \vec{\sigma_{\rm{diff}}}.
    \label{EQ:AN_approx}
\end{equation}
Here $\vec{\sigma_{\rm{diff}}} = \vec{\sigma} (\vec{0})$ refers to the eigenvalues after a brief $\Delta t$ period of diffusion. 
$\Sigma_{\rm{adv}} \in \mathbb{R}^{N \times N}$ is the matrix of eigenvalues after $N$ periods of advection (one for each mode).
% The $k$th row of $J_0$ is then given by the gradient of the $k$th mode's eigenvalue $\nabla \sigma_k$.
In this way, an arbitrary element of $\sigma_{\rm{adv}}$ represents how a particular mode's eigenvalue responds to the advection of another mode. 
We then approximate the Jacobian matrix of $\vec{\sigma}(\vec{A^2})$ via first-order finite differences at the point of our coarse estimate, requiring $N^2$ more individual EVP solves.
After its construction, the Jacobian is updated with each guess according to Broyden's method for root-finding in multi-dimensional functions \cite{Broyden}.
We find $\vec{\sigma}(\vec{A^2})$ does indeed have a unique root provided the time step is not too large and there are no numerical instabilities, as outlined in appendix \ref{sec:timestep}.
Presumably this is due to the dependence of $W(z)$ and $\theta(z)$ on $\bar{T}$.
Over the course of a large time step, $\bar{T}$ evolves according to \eq{EQ:T0_IVP} and the eigenfunctions might cease to provide a stabilizing influence.

>>>>

\subsection{Treatment of multiple marginally-stable modes} \label{sec:multiple_modes}
In most cases, we eventually encounter eigenvalue spectra with multiple simultaneously marginal modes.
If ignored, we are forced to reduce the timestep and allow the modes to alternate which is marginally stable.
Instead, we generalize the advective term in \eq{EQ:T0_IVP} to accommodate $N$ simultaneously marginal modes
\begin{equation}
    \langle w' T' \rangle_x = \sum_{n = 1}^{N} 4 A_n^2 \int \, \Re\left[ W_n \theta_n^* \right] \,dz
\end{equation}
where $W_n$ and $\theta_n$ are the eigenmodes associated with $k_x = \frac{n\pi}{2}$. 
There are now $N$ amplitudes to solve for and $N$ eigenvalues to keep marginally-stable. 
Given a small fixed time step $\Delta t$, let $\vec{A^2}$ be the amplitude vector and $\vec{\sigma}(\vec{A^2})$ be the dependent eigenvalues.
We expect a function $\vec{\sigma} \, : \, \mathbb{R}^N \to  \mathbb{R}^N$ to have isolated roots $\vec{A^2_m}$ (should they exist). 
A coarse approximation $\vec{A^2_0}$ for the appropriate amplitude vector $\vec{A^2_m}$ is given by a generalized form of \eq{EQ:amp_approx}:
\begin{equation}
    \vec{A^2_0} = -\Sigma_{\rm{adv}}^{-1} \vec{\sigma_{\rm{diff}}}.
    \label{EQ:AN_approx}
\end{equation}
Here $\vec{\sigma_{\rm{diff}}} = \vec{\sigma} (\vec{0})$ refers to the eigenvalues after a brief $\Delta t$ period of diffusion. 
$\Sigma_{\rm{adv}} \in \mathbb{R}^{N \times N}$ is the matrix of eigenvalues after $N$ periods of advection (one for each mode).
% The $k$th row of $J_0$ is then given by the gradient of the $k$th mode's eigenvalue $\nabla \sigma_k$.
In this way, an arbitrary element of $\Sigma_{\rm{adv}}$ represents how a particular mode's eigenvalue responds to the advection of another mode. 
At the point of our coarse estimate $\vec{A^2_0}$, we approximate the Jacobian matrix 
\begin{equation}
    J = \begin{bmatrix}
        \nabla \sigma_1 (A_1, A_2, ..., A_N) \\
        \nabla \sigma_2 (A_1, A_2, ..., A_N) \\
        \vdots \\
        \nabla \sigma_N (A_1, A_2, ..., A_N) 
    \end{bmatrix}
\end{equation}
via first-order finite differences, requiring $N^2$ more individual EVP solves.
A refined estimate is then given by
\begin{equation}
    \vec{A^2_1} = -J^{-1} \vec{\sigma}(\vec{A^2_0}).
\end{equation}
Once constructed we use each subsequent guess to update the Jacobian iteratively according to Broyden's method for root-finding in multi-dimensional functions \cite{Broyden}.
We find $\vec{\sigma}(\vec{A^2})$ does indeed have a unique root provided the time step is not too large and there are no numerical instabilities, as outlined in appendix \ref{sec:timestep}.
Presumably this is due to the coupling of $W(z)$ and $\theta(z)$ with $\bar{T}$.
Over the course of a large time step, $\bar{T}$ evolves according to \eq{EQ:T0_IVP} and the eigenfunctions might cease to provide a stabilizing influence.
